---
title: "Revisiting Knudson's 2 Hit Hypothesis: A Lesson in Data Reproducibility"
author: "Daniel Pique"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{knudson}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Knudson's ["Mutation and cancer: statistical study of retinoblastoma"](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC389051/) (1971) forms the basis for the 2-hit hypothesis in cancer and has 8000+ citations in the biomedical literature. Here, I attempt to reproduce [Figure 1](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC389051/?page=4) from this paper using the data provided in [Table 1](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC389051/?page=2), which is also provided within this package as the object `k`. 


## Load the `knudson` library and look at Table 1

We'll also load some other packages that we'll need for the rest of the analysis. 

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning=FALSE,
  error=FALSE, 
  message=FALSE
)
```


```{r warning=FALSE, error=FALSE, message=FALSE, fig.height = 4, fig.width = 7}
library(knudson)
library(tidyverse)
library(patchwork)
library(png)
library(knitr)
library(kableExtra)


t1_png <- png::readPNG("knudson_tab1_1971_PNAS.png")

t1_png_gg <- ggplot() +  
  annotation_custom(grid::rasterGrob(t1_png, 
                               width = unit(1,"npc"), 
                               height = unit(1,"npc")), 
                    -Inf, Inf, -Inf, Inf)
t1_png_gg

# `k` is a machine-readable version of Table 1 
kable(knudson::k) %>%
  kable_styling(position="center") %>%
  scroll_box(width = "100%", height = "200px")

```


Now, let's take a look at figure 1 and compare it with the data in Table 1.

```{r, echo = TRUE, fig.height = 6, fig.width = 4.5}
knud_png <- png::readPNG("knudson_fig1_1971_PNAS.png")

fig1_orig <- ggplot() +  
  annotation_custom(grid::rasterGrob(knud_png, 
                               width = unit(1,"npc"), 
                               height = unit(1,"npc")), 
                    -Inf, Inf, -Inf, Inf)

fig1_orig
```

Of note, it seems that the x-axis in Figure 1 ends at 60 months. However, there are 2 patients in Table 1 who are diagnosed at or after age 60 months. To make sure, let's count the number of data points represented in this figure.

```{r, fig.height = 6, fig.width = 4.5}
knud_png_annot <- png::readPNG("knudson_fig1_1971_PNAS_annot.png")

knud_fig1_annot <- ggplot() +  
  annotation_custom(grid::rasterGrob(knud_png_annot, 
                               width = unit(1,"npc"), 
                               height = unit(1,"npc")), 
                    -Inf, Inf, -Inf, Inf)
knud_fig1_annot
```

We observe that there are fewer data points than are listed in the legend (20 unilateral and 17 bilateral cases compared with the expected 25 and 23, respectively).

Why are there fewer points in the plot than expected? The y-axis reports the "fraction of cases not yet diagnosed at the age indicated." Therefore, it is possible that the author collapsed cases occuring in the same month into a single data point. 

We'll recreate figure 1 with this approach in mind.


## Recreating Figure 1

```{r, fig.height = 6, fig.width = 8}

k2 <- k %>%
  arrange(side, -age_at_dx) %>%
  group_by(side) %>%
  mutate(idx_sum = 1 / length(side), 
         prop_of_tot = cumsum(idx_sum) - idx_sum) %>%
  ungroup %>%
  mutate(side=factor(side, levels=c("unilat", "bilat"))) %>%
  group_by(side, age_at_dx) %>%
  ## collapse cases occuring in the same month into a single data point
  filter(prop_of_tot == min(prop_of_tot)) 

y_axis_seq <- c(0.01, seq(0.02, 0.1, 0.02), seq(0.2, 1, 0.2))
y_axis_seq_char = format(y_axis_seq, digits=2)

fig1 <- ggplot(k2, aes(x=age_at_dx, y = prop_of_tot, group=side, shape=side)) +
  geom_point() +
  scale_shape_manual(values=c(21,22), 
                     labels=c("Unilateral Cases (25)","Bilateral Cases (23)")) + 
  scale_y_log10(breaks = y_axis_seq, 
                limits = c(0.01, 1.04), 
                labels = as.character(gsub("^0\\.", "\\.", y_axis_seq_char)),
                expand = c(0, 0)) +
  scale_x_continuous(breaks = seq(0,60, 10), 
                     limits = c(-0.1, 75), 
                     expand = c(0, 0)) +
  xlab("AGE IN MONTHS") + 
  ylab("FRACTION OF CASES NOT YET DIAGNOSED AT AGE INDICATED") +
  theme_classic() + 
  theme(aspect.ratio = 1.5, 
        legend.position = c(0.3,0.16), 
        legend.title=element_blank()) +
  ggtitle("Reproduced in R") + geom_abline(intercept = 0, slope = -1/30, color="black", 
                 linetype="solid", size=0.5) + 
  stat_function(fun = function(t) -4 * 10^{-4} *t^2, linetype="longdash", color="black")

k2_missing <- k2 %>% filter(age_at_dx >= 60 | {age_at_dx == 47 & side == "unilat"})

fig1_missing <- fig1 + 
  geom_point(data = k2_missing, aes(x=age_at_dx, y = prop_of_tot), pch=21, fill=NA, color = "red", size=4)


fig1_missing + fig1_orig + plot_layout(widths = c(1, 1.35))
```


## What happens when we compare this figure to the original figure?

We notice that there are 3 datapoints (red circles) missing from Figure 1. Two datapoints from patients age >=60 are missing, and a third datapoint for unilateral retinoblastoma patient 08997, age 47 months, is also missing. 

```{r, fig.height = 6, fig.width = 8}

k2_missing <- k2 %>% filter(age_at_dx >= 60 | {age_at_dx == 47 & side == "unilat"})

fig1_missing <- fig1 + 
  geom_rect(mapping = aes(xmin=0.1, xmax=60, ymin=0.01, ymax=1), fill=NA, color="red", linetype = 2) + 
  geom_point(data = k2_missing, aes(x=age_at_dx, y = prop_of_tot), pch=21, fill=NA, color = "red", size=4)

fig1_missing + 
  fig1_orig + 
  plot_layout(widths = c(1, 1.35))

```


We also produce a plot with each point annotated, and compare this to a manually annotated version of the original figure:
```{r fig.height = 6, fig.width = 8}
k3 <- k2 %>% 
  ungroup %>% 
  group_by(side) %>% 
  arrange(age_at_dx) %>% 
  mutate(idx = 1:n())

fig1_annot <- fig1 +
  geom_text(data=k3, aes(label = idx), size=3, nudge_x = 2) +   
  scale_x_continuous(breaks = seq(0,60, 10), 
                     limits = c(-0.1, 80), 
                     expand = c(0, 0))
  
fig1_annot + 
  knud_fig1_annot + 
  plot_layout(widths = c(1, 1.35))

```


We note that the curve $logS = -4 \times 10^{-4}t^2$ is a better fit than the curve $logS = -4 * 10^{-5} *t^2$ reported in the paper, suggesting that the exponent reported in the paper is off by an order of magnitude. 


## Conclusion

This vignette highlights the importance of data reproducibility in the scientific literature.

We also notice that this paper was ['communicated'](https://www.pnas.org/content/106/37/15518) to PNAS, meaning that it was 'believed... (by a NAS member)... to be of sufficient significance to warrant publication.' This means that it did not undergo the standard rigorous process of peer review that generally involves 2-3 field experts.

