---
title: "knudson"
author: "Daniel Pique"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{knudson}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Knudson's "Mutation and cancer: statistical study of retinoblastoma" (1971) has 7900+ citations in the biomedical literature. Here, I recreate the lone figure in this paper and release a package containing the data from table 1 in this paper. The I show why statistics and reproducibility needs to be a part of the curriculum in medical schools and graduate schools.

The purpose of this package is to provide data from 2 seminal papers published by Alfred Knudson. The first, entitled "Mutation and Cancer: Statistical Study of Retinoblastoma," was published in 1971 in PNAS. This paper has 7900+ citations and forms the basis for the 2-hit hypothesis in cancer. The data from table 1 of this paper is stored in the object `k`. 

A second paper, entitled "Mutation and Cancer: A Model for Wilms' Tumor of the Kidney," followed shortly thereafter in 1972 and confirmed the findings from the 1971 paper. The authors were Alfred G. Knudson and Louise Strons. The data from Table 1 in this paper is stored in the object `w`.

We notice that this paper was ['communicated'](https://www.pnas.org/content/106/37/15518) to PNAS, meaning that it was 'believed... (by a NAS member)... to be of sufficient significance to warrant publication.' This means that it did not undergo the standard rigorous process of peer review that generally involves 2-3 field experts.



```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


```{r}
library(knudson)
library(tidyverse)
head(knudson::k)
head(knudson::w)

ggplot(k, aes(x=side, y= age_at_dx)) + geom_violin() + geom_jitter(alpha=0.5, width=0.1)

k2 <- k %>%
  arrange(side, -age_at_dx) %>%
  group_by(side) %>%
  mutate(idx_sum = 1 / length(side)) %>%
  mutate(prop_of_tot = cumsum(idx_sum))

y_axis_seq <- c(0.01, seq(0.02, 0.1, 0.02), seq(0.2, 1, 0.2))

two_hit_fxn <- function(y){
  abs(log(y, 10)/(-4*10^(-4))) %>% sqrt
}

log(.2, 10) * -30

(log(.2, 10) / (-4*10^(-4)) ) %>% sqrt

ggplot(k2, aes(x=age_at_dx, y = prop_of_tot, group=side, shape=side)) +
  
  geom_abline(intercept = 0, slope = -1/30, color="black", 
                 linetype="solid", size=0.5) + # can set x limit to 50 here 
  stat_function(fun = function(t) -4 * 10^{-5} *t^2, linetype="longdash", color="red") + #raise to the 10^ power 

  stat_function(fun = function(t) -5 * 10^{-4} *t^2, linetype="longdash") + #raise to the 10^ power, incorrect from what's listed 
    stat_function(fun = function(t) -4 * 10^{-4} *t^2, linetype="longdash", color="green") + #raise to the 10^ power, incorrect from what's listed 
      stat_function(fun = function(t) -3 * 10^{-4} *t^2, linetype="longdash", color="blue") + #raise to the 10^ power, incorrect from what's listed 
      stat_function(fun = function(t) -3.5 * 10^{-4} *t^2, linetype="longdash", color="purple") + #raise to the 10^ power, incorrect from what's listed 


  
    stat_function(fun = function(t) 2.714^(-4*10^{-5} * t^2), linetype="longdash", color="green") + #raise to the 10^ power, try the log version 
  geom_hline(yintercept = .1) +

  
  geom_point() +
  scale_shape_manual(values=c(22,21)) + 
  scale_y_log10(breaks= y_axis_seq, 
                limits = c(0.01, 1), 
                labels = as.character(gsub("^0\\.", "\\.", y_axis_seq)),
                expand = c(0, 0)) +
  scale_x_continuous(breaks = seq(0,60, 10), 
                     limits = c(-0.1, 75), 
                     expand = c(0, 0)) +
  xlab("Age in months") + 
  ylab("Fract. of cases not yet diagnosed at age indicated") +
  theme_classic()

#calculate the best fit line

#add the two curves
logS = -t/30 # one hit curve
logS = -4 * 10^{-5} *t^2 #
logS = -5 * 10^{-4} * t^2 #-5 * 10^{-4} * t^2

```

The original figure is included here:
[figure 1 knudson 1971 PNAS](img/knudson_fig1_PNAS.png)



We note two differences. The first is that the curve $logS = -5 \times 10^{-4}t^2$ is a better fit than the curve $logS = -4 * 10^{-5} *t^2$ reported in the paper, suggesting that these numbers may have been flipped. The second is that knudson reports the use of the natural logarithm, $ln S = -kt^2$, when in fact the graphic shows the use of $log_{10}$.


Below, we quantitate the goodness of fit from these different models.
```{r}

f_log10_2 <- formula(prop_of_tot ~ I(10^(-age_at_dx*0.02607)))

f_log10_2 <- formula(log(prop_of_tot,10) ~ -age_at_dx/30)

#
f_log10_2 <- formula(prop_of_tot ~ I(10^(-age_at_dx/30)))# logS = -t/30 prop_of_tot ~ 10^-4 * 10^{-5} age_at_dx*^2))
optim_preds_knud <- predict.lm(object = f_log10_2, newdata = k2.bilat) # %>% {10^.}

preds <- model.frame(formula = f_log10_2, data = k2, subset = side == "bilat") %>% 
  as_tibble() %>% 
  mutate(resid = prop_of_tot - `I(10^(-age_at_dx/30))`) #`I(10^(-age_at_dx * 0.02607))`) #
hist(preds$resid)
rss <- sum(preds$resid^2)
rss #knudson's rss
rss2 <- sum(abs(preds$resid))
rss2


#now, we look at what an optimal rss would be
#optim_mod <- lm(formula = prop_of_tot ~ I(10^-(age_at_dx/30)), data = k2, subset = side == "bilat")
k2.bilat <- k2 %>% filter(side == "bilat")

#optim_mod <- lm(formula = I(log(prop_of_tot, 10)) ~ 0 + age_at_dx, data = k2, subset = side == "bilat")
#optim_mod2 <- lm(formula = k2.bilat$prop_of_tot ~ k2.bilat$age_at_dx), data = k2, subset = side == "bilat")

optim_mod <- lm(formula = prop_of_tot ~ log(age_at_dx,10), data = k2.bilat)


optim_mod_int_0 <- lm(formula = log(prop_of_tot,10) ~ age_at_dx + 0, data = k2.bilat)

sum(optim_mod_int_0$residuals^2) #sum of sq

optim_mod <- lm(formula = log(prop_of_tot,10) ~ age_at_dx, data = k2.bilat)
sum(optim_mod$residuals^2) #sum of sq

optim_mod <- lm(formula = prop_of_tot ~ age_at_dx, data = k2.bilat)


# now get the prediction from the model
optim_preds <- predict.lm(object = optim_mod, 
                          k2.bilat) #%>% {10^.}
                            #mutate(prop_of_tot = log(prop_of_tot,10)))  %>% {10^.}
# now get the observed from the data
k2.bilat$prop_of_tot

(optim_preds - k2.bilat$prop_of_tot)^2 %>% sum
#look up RSS on log scale

summary(residuals(optim_mod))
sum(optim_mod$residuals^2)


preds2 <- model.frame(formula = optim_mod, data = k2.bilat) %>%
  as_tibble() %>% 
  mutate(prop_of_tot_pred = optim_preds, #10^`log(prop_of_tot, 10)`,
         prop_of_tot_obs = k2.bilat$prop_of_tot,
         resids = prop_of_tot_pred - prop_of_tot_obs)
hist(preds2$resids)
rss <- sum(preds2$resids^2)
rss #knudson's rss

ggplot(preds2, aes(x=age_at_dx, y=prop_of_tot_obs)) + 
         geom_point() +
      # stat_function(fun = #optim_mod, 
      #               function(t) .76077 + -0.0723*t,
      #               linetype="longdash", color="purple") + 
        stat_function(fun = #optim_mod, 
                    function(t)  -0.02607*t,
                    linetype="longdash", color="blue") +
          stat_function(fun = #optim_mod, 
                    function(t) -1/30*t,
                    linetype="longdash", color="green") +
          stat_function(fun = 
                    function(t) 1.13 + -0.65177*t,
                    linetype="longdash", color="orange") +  

  scale_y_log10() #raise to the 10^ power, incorrect from what's listed 

hist(optim_mod$residuals)

sum(optim_mod$residuals^2)
optim_preds <- predict.lm(object = optim_mod, k2)#model.frame(formula = )

plot(formula = log(prop_of_tot, 10) ~ age_at_dx, data = k2)
predict(optim_mod, newdata = k2)



f_log10 <- formula(prop_of_tot ~ 10^-4 * 10^{-5} age_at_dx*^2))
model.frame(formula = f_log10, data = k2)
k2
logS = -4 * 10^{-5} *t^2 #
logS = -5 * 10^{-4} * t^2 #-5 * 10^{-4} * t^2

```


Create a test prediction with some synthetic data
see if I can predict the model

```{r}
data ()
```



Now, we recreate Figure 1 from Knudson and Strons, 1972 
```{r}
w2 <- w %>%
  arrange(side, -age_at_dx) %>%
  group_by(side) %>%
  mutate(idx_sum = 1 / length(side)) %>%
  mutate(prop_of_tot = cumsum(idx_sum))

```

